---
title: "Population Pharmacokinetic Modeling (PopPK) using R"
output: html_notebook
---

# INTRODUCTION 
110 patients received a 100 mg oral dose of Drug X. Plasma concentrations were sampled from 1 to 10 hours for each patient. Sex and age along with concentration data are provided (0 = female, 1 = male). 

## Objective
Construct a population PK model with covariates. 

## Setup
```{r warning=F, message=F}
#install.packages("saemix")

library(tidyverse)
library(saemix)
library(PKNCA)
library(trelliscopejs)
library(ggsignif)

```



# EXPLORATORY ANALYSIS 


## Data Import 
```{r}
all.data <- read.csv("Data files/PopPK Data.csv") %>% rename("Dose" = "Aa")
all.data$id <- as.factor(all.data$id)
head(all.data)
```


## Concentration vs Time Plot 
```{r}
CT1.plot <- ggplot(data = all.data, aes(x = Time, y = CObs))+ 
  geom_line(aes(group = id, color = id))+ 
  geom_point(aes(color = id))+
  lims(x = c(0,10))+
  theme(legend.position = "none")
  
CT1.plot
CT1.plot + scale_y_log10(limits = c(0.001,10))

```


## Concentration vs Time by Covariates 
```{r}
CT1.plot + scale_y_log10(limits = c(0.001,10)) + facet_grid(.~Gender)
```


# INITIAL ESTIMATES

## Non-Compartmental Analysis (NCA)
```{r message=F}
naive.conc <- all.data %>% 
  group_by(Time) %>% 
  summarize(Conc = mean(CObs), id = "All Subjects") %>% 
  ungroup()

naive.conc$Conc[naive.conc$Time == 0] <- 0

naive.dose <- tibble(Dose = 100, Time = 0, id = "All Subjects")

interval.data <- tibble(start = 0, 
                        end = 10,
                        aucall = T,
                        auclast = T, 
                        aucinf.obs = T,
                        aucinf.pred = T,
                        cmax = T, 
                        vz.obs = T,
                        cl.obs = T)


dose.data <- PKNCAdose(naive.dose, Dose~Time|id)
conc.data <- PKNCAconc(naive.conc, Conc~Time|id)

NCA.data <- PKNCAdata(conc.data, intervals = interval.data)
NCA.result <- pk.nca(NCA.data)
NCA.summary <- NCA.result$result %>%
  spread(key = PPTESTCD, value = PPORRES) %>% 
  mutate(CL = naive.dose$Dose[1]/aucinf.obs,
         V = CL/lambda.z) %>%
  select(aucinf.obs, half.life, cmax, CL, V, lambda.z) %>% 
  gather(1:6, key = "Parameter", value = "Value")

ggplot(naive.conc, aes(x = Time, y = Conc))+
  geom_line()+
  geom_point()+
  scale_y_log10()

```


### PK Parameter Summary 
```{r}
NCA.summary
```



# BASE MODEL 

## Model Structure
```{r}
pk.model1 <- function(psi, t){
  D  <- 100
  ka <- psi[1]
  V  <- psi[2]
  ke <- psi[3]
  f  <- D*ka/V/(ka-ke)*(exp(-ke*t)-exp(-ka*t)) 
  return(f)}

```


## Parameter Optimization

### Naive Data Estimate using NLS
```{r}
pkm1 <- nls(CObs ~ pk.model1(psi = psi, t = Time), 
            start=list(psi=c(ka=1, V=66.69, ke=0.126)), 
            data=all.data)
coef(pkm1)

grid <- tibble(Time = seq(0,11,0.1))
grid$pred <- predict(pkm1, grid)

ggplot(all.data, aes(x = Time, y = CObs))+
  geom_point()+
  geom_point(color = "red", size = 4, data = naive.conc, aes(x = Time, y = Conc))+
  geom_line(data = grid, aes(x = Time, y = pred), size = 1, color = "red")+ lims(x = c(0,12))

```

### SAEM Optimization
```{r}
model1cpt <- function(psi,id,x) { 
  D   <- 100
  t   <-x[,1] 
  ka  <-psi[id,1]
  V   <-psi[id,2]
  ke  <-psi[id,3]
  fpred <-D*ka/(V*(ka-ke))*(exp(-ke*t)-exp(-ka*t))
  return(fpred)
  } 
```

```{r echo=F}

saemix.data <- saemixData(name.data = all.data %>%
                            filter(Time !=0),
                          name.group      = "id",
                          name.predictors = "Time",
                          name.response   = "CObs")

saemix.model <- saemixModel(model = model1cpt, 
                            psi0  = c(ka=4.2,V=78.89,ke = 0.0834))

saemix.options <- list(map=TRUE, fim=TRUE, ll.is=FALSE, 
                       seed=23456, print = F, save = F, save.graphs = F)
saemix.fit1    <- saemix(saemix.model, saemix.data, saemix.options)

```


```{r}
saemix.fit1@results

psi <- psi(saemix.fit1) %>% mutate(id = seq(1,110,1))
psi$Gender <- all.data$Gender[all.data$Time == 0]
psi$Age <- all.data$Age[all.data$Time == 0]
psi <- psi %>% select(id,ka,V,ke,Gender,Age)



```

```{r message = F}

grid1 <- grid %>% mutate(SAEM_pred = 100*1.238/(101.36*(1.238-0.5))*
                  (exp(-0.5*Time)-exp(-1.238*Time))) %>% 
  gather(2:3, key = "Method", value = "Pred") %>% 
  mutate(Method = if_else(Method == "pred","NLS","SAEM"))

ggplot(data = grid1 %>% filter(Method == "SAEM"), aes(x = Time, y = Pred))+
  geom_line(aes(color = Method), size = 1)+
  geom_point(data = all.data, aes(x = Time, y = CObs), 
             size = 3, shape = 21, color = "blue")
```


## Additive Error Model Evaluation
#### DV vs IPRED
```{r}
saemix.plot.obsvspred(saemix.fit1, level = 1)
```


#### DV.IPRED vs IVAR
```{r message=F}
pred_1cmt <- function(ka, V, ke, Time){
     C <- 100*ka/(V*(ka-ke))*(exp(-ke*Time)-exp(-ka*Time))
  return(C)
}

grid2 <- psi %>% 
  expand_grid(Time = seq(0,11, 0.1)) %>% 
  mutate(SAEM_pred = pred_1cmt(ka, V, ke, Time)) %>% 
  mutate(id = as_factor(id)) %>% left_join(all.data)



PopPK.plot1 <- ggplot(data = grid2, aes(x = Time, y = CObs, group = id))+
  geom_line(size = 1, aes(x = Time, y = SAEM_pred, color = id))+
  geom_point(shape = 21, size = 3, aes(fill = id), alpha = 0.5)+
  theme(legend.position = "none")

PopPK.plot1
```

#### CWRES vs IVAR & IPRED
```{r fig.asp=1}
saemix.plot.scatterresiduals(saemix.fit1, level = 1)
saemix.plot.distribresiduals(saemix.fit1, level = 1)
```

#### Final Parameters vs Initial Estimates

## Proportional Error Model Evaluation ----

```{r}
saemix.model.prop <- saemixModel(model = model1cpt, 
                            psi0  = c(ka=1.238,V=101.361,ke=0.498), 
                            error.model = "proportional")

saemix.fit2    <- saemix(saemix.model.prop, saemix.data, saemix.options)

```


```{r}
saemix.fit2@results

psi.prop <- psi(saemix.fit2) %>% mutate(id = seq(1,110,1))
psi.prop$Gender <- all.data$Gender[all.data$Time == 0]
psi.prop$Age <- all.data$Age[all.data$Time == 0]
psi.prop <- psi.prop %>% select(id,ka,V,ke,Gender,Age)

psi.prop

eta.prop0 <- matrix(eta(saemix.fit2, type = "mean"), 
                    ncol = 3, nrow = 110, 
                    dimnames = list(seq(1,110,1), c("ka","V","ke")))

eta.prop <- as_tibble(eta.prop0) %>% mutate(id = seq(1,110,1))
eta.prop$Gender <- all.data$Gender[all.data$Time == 0]
eta.prop$Age <- all.data$Age[all.data$Time == 0]
eta.prop <- eta.prop %>% select(id,ka,V,ke,Gender,Age)

eta.prop

```

#### DV vs IPRED
#### DV.IPRED vs IVAR
```{r}
grid.prop <- psi.prop %>% 
  expand_grid(Time = seq(0,11, 0.1)) %>% 
  mutate(SAEM_pred = pred_1cmt(ka = ka, V = V, ke = ke, Time = Time)) %>% 
  mutate(id = as_factor(id)) %>% left_join(all.data)



PopPK.plot.prop <- ggplot(data = grid.prop, aes(x = Time, y = CObs, group = id))+
  geom_line(size = 1, aes(x = Time, y = SAEM_pred, color = id))+
  geom_point(shape = 21, size = 3, aes(fill = id), alpha = 0.5)+
  theme(legend.position = "none")

PopPK.plot.prop+ scale_y_log10()
```

#### CWRES vs IVAR
```{r fig.asp=1}
saemix.plot.scatterresiduals(saemix.fit2, level = 1)
saemix.plot.distribresiduals(saemix.fit2, level = 1)
```

## Model Comparison 
### AIC & BIC
```{r}
compare.saemix(saemix.fit1, saemix.fit2, method = "lin")

```

### Visual Predictive Checks


#### Add + Prop Error Model
```{r}
fit2.coef <- coef.SaemixObject(saemix.fit2)$fixed

saemix.model.comb <- saemixModel(model = model1cpt, 
                            psi0  =fit2.coef, 
                            error.model = "combined", verbose = F)

saemix.fit3    <- saemix(saemix.model.comb, saemix.data, saemix.options)

saemix.fit3@results

```

#### CWRES vs IVAR 

# COVARIATE MODEL

```{r fig.asp=0.9}

psi.covar <- psi.prop %>% 
  mutate(Gender = as_factor(if_else(Gender == 0, "Female", "Male"))) %>% 
  gather(c(ka:ke), key = "Parameter", value = "Value")

eta.covar <- eta.prop %>% 
  mutate(Gender = as_factor(if_else(Gender == 0, "Female", "Male"))) %>% 
  gather(c(ka:ke), key = "Parameter", value = "Value")

ggplot(psi.covar, aes(x = Gender, y = Value))+
  geom_boxplot()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  stat_summary(geom = "crossbar", aes(color = Gender), fun = mean)+
  stat_summary(geom = "point", size = 3, shape = 21, 
               aes(fill = Gender), fun = mean)+
  geom_signif(comparisons = list(c("Male","Female")), 
              test = "t.test", test.args = "two.sided", 
              map_signif_level = c("***" = 0.001, "**"=0.01, "*"=0.05))


ggplot(eta.covar, aes(x = Gender, y = Value))+
  geom_boxplot()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  stat_summary(geom = "crossbar", aes(color = Gender), fun = mean)+
  stat_summary(geom = "point", size = 3, shape = 21, 
               aes(fill = Gender), fun = mean)+
  geom_signif(comparisons = list(c("Male","Female")), 
              test = "t.test", test.args = "two.sided", 
              map_signif_level = c("***" = 0.001, "**"=0.01, "*"=0.05))

# Age vs Parameters
ggplot(psi.covar, aes(x = Age, y = Value))+
  geom_point()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  geom_smooth(method = "glm")

ggplot(eta.covar, aes(x = Age, y = Value))+
  geom_point()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  geom_smooth(method = "glm")


# Interaction between Gender and Age
ggplot(psi.covar, aes(x = Age, y = Value))+
  geom_point()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  geom_smooth(method = "glm", aes(color = Gender))

ggplot(eta.covar, aes(x = Age, y = Value))+
  geom_point()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  geom_smooth(method = "glm", aes(color = Gender))

psi.corr <- psi.prop %>% select(-id, -Gender, -Age) %>% plot()
eta.corr <- eta.prop %>% select(-id, -Gender, -Age) %>% plot()

```


```{r }

saemix_covar.data <- saemixData(name.data = all.data %>%
                            filter(Time !=0),
                          name.group      = "id",
                          name.predictors = "Time",
                          name.response   = "CObs", 
                          name.covariates = c("Gender","Age"))

saemix_covar_model.prop <- saemixModel(model = model1cpt, 
                            psi0  = fit2.coef, 
                            error.model = "proportional",
                            covariate.model = matrix(c(1,1,1,0,0,0), 
                                                     nrow = 2, ncol = 3, 
                                                     byrow = T))

saemix.options <- list(map=TRUE, fim=TRUE, ll.is=FALSE, 
                       seed=23456, print = F, save = F, save.graphs = F)
saemix_covar.fit1    <- saemix(saemix_covar_model.prop, 
                               saemix_covar.data, 
                               saemix.options)
```

## Stepwise Covariate Search
```{r}

saemix_covar.fit1@results
# res.stepwise <- step.saemix(saemix_covar.fit1, direction="both", 
#                             covariate.init = matrix(c(1,0,1,0,1,1), 
#                                                     nrow = 2, ncol = 3, 
#                                                     byrow = T),
#                             trace = T)

res.stepwise
```

## Covariate Plots (w/best base model) 
### ETA vs Covariate
```{r}
eta.covar0 <- matrix(eta(saemix_covar.fit1, type = "mean"), 
                    ncol = 3, nrow = 110, 
                    dimnames = list(seq(1,110,1), c("ka","V","ke")))

eta.covar <- as_tibble(eta.covar0) %>% mutate(id = seq(1,110,1))
eta.covar$Gender <- all.data$Gender[all.data$Time == 0]
eta.covar$Age <- all.data$Age[all.data$Time == 0]
eta.covar <- eta.covar %>% select(id,ka,V,ke,Gender,Age)

eta.covar

eta.covar <- eta.covar %>% 
  mutate(Gender = as_factor(if_else(Gender == 0, "Female", "Male"))) %>% 
  gather(c(ka:ke), key = "Parameter", value = "Value")

ggplot(eta.covar, aes(x = Gender, y = Value))+
  geom_boxplot()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")+
  stat_summary(geom = "crossbar", aes(color = Gender), fun = mean)+
  stat_summary(geom = "point", size = 3, shape = 21,
               aes(fill = Gender), fun = mean)+
  geom_signif(comparisons = list(c("Male","Female")),
              test = "t.test", test.args = "two.sided",
              map_signif_level = c("***" = 0.001, "**"=0.01, "*"=0.05))

ggplot(eta.covar, aes(x = Age, y = Value))+
  geom_point()+
  geom_smooth()+
  facet_wrap(.~Parameter, ncol = 2, scales = "free_y", strip.position = "top")

```

```{r}
saemix.plot.scatterresiduals(saemix_covar.fit1, level = 1)
saemix.plot.distribresiduals(saemix_covar.fit1, level = 1)
```

## Parameter Estimates of Covariate Model
## Data Visualization of Covariate Model


# COVARIANCE MODEL

## Diagonal Model
### ETA Scatter Plots (Theta Correlations)
## Non-Diagonal Model (Based on Evidence)
## Covariance Model Comparison


# MODEL QUALIFICATION 

## Boostrap - Quality of Theta Estimates
## Bootstrap vs Final Model Estimates
## Visual Predictive Check (Prediction Intervals)


# MODELING AND SIMULATION 

## Single Dose
## Ascending Dose
## Multiple Dosing Scenario
## Multiple Ascending Dose